---
title: "drop_sa"
output: html_document
date: "2022-08-18"
---

```{r setup, include=FALSE}
library(ComplexHeatmap)
library(tidyverse)

theme_set(theme_bw())

knitr::opts_chunk$set(echo = TRUE)
```

```{r import_fraser}

fraser_results_all.tib <- read_tsv(here::here("data", "trios_star_results.tsv.gz"))

fraser_results.tib <- fraser_results_all.tib %>%
    filter(padjust < 0.1) %>%
    mutate(DROP_GROUP = case_when(str_detect(sampleID, "^DH") ~ "fraser,samples_bcchr",
                                  str_detect(sampleID, "^(HG|NA)") ~ "fraser,controls_geuvadis",
                                  str_detect(sampleID, "^GTEX") ~ "fraser,controls_gtex")) %>%
    separate(DROP_GROUP, into = c("DROP_GROUP_RULE", "DROP_GROUP_ORIGIN"), sep = ",")

fraser_results.tib %>%
    # filter(hgncSymbol == "PHF21A")
    filter(sampleID == "DH242401") %>%
    arrange(padjust)

```

```{r plot_distribution_by_region, fig.height=8.6, fig.width=8.6}

seqnames.order <-
    c(paste0("chr", seq(1, 22)),
      "chrX",
      "chrM")

# summarise junctions by sample ID and chromosome
fraser_results_by_chrom.tib <-
    fraser_results.tib %>%
    filter(DROP_GROUP_ORIGIN == "samples_bcchr") %>%
    group_by(sampleID, seqnames, type) %>%
    tally() %>%
    mutate(seqnames = factor(seqnames, levels = seqnames.order),
           family_id = map_chr(sampleID,     ~ str_match(., "(DH[0-9]{4})([0-9]{2}|HC)")[2]),
           individual_id = map_chr(sampleID, ~ str_match(., "(DH[0-9]{4})([0-9]{2}|HC)")[3]))

ggplot(fraser_results_by_chrom.tib) +
    geom_point(aes(x = seqnames, y = sampleID, size = n, color = type),
               alpha = 0.5) +
    facet_grid(rows = vars(family_id), scales = "free_y") +
    theme(axis.text.x.bottom = element_text(angle = 45, hjust = 1))

ggsave(here::here("plots", "fraser_results_by_chr_and_type.png"),
       width = 8, height = 8)

```

## Quantify Gene Hit Frequency

Want to count mutations by affected gene to identify hotspot genes with common splice defects in the population to select against them.

```{r count_gene_hit_freq}

fraser_results_by_gene.tib <-
    fraser_results.tib %>%
    group_by(DROP_GROUP_ORIGIN, hgncSymbol) %>%
    tally() %>%
    pivot_wider(names_from = DROP_GROUP_ORIGIN, values_from = n) %>%
    mutate(across(-hgncSymbol, replace_na, 0)) %>%
    arrange(desc(samples_bcchr))

fraser_results_by_gene.tib

```

## Identify Unique Splice Defects in Patient Samples

Will first try to look at distribution of hits and see if we can filter against everything that's observed in the population first.

```{r filter_population_hits}

fraser_sample_hits <-
    fraser_results_by_gene.tib %>%
    filter(controls_gtex <= 5)  # controls_geuvadis == 0, 

fraser_sample_hits

```

This leaves 86 hits, will pull them out of the results to check significance and which individuals are affected. Will visualize as a heatmap.

```{r extract_hits_from_sample_results, fig.width=8.6, fig.height=14}

fraser_results_sample_hits_table <-
    fraser_results.tib %>%
    filter(DROP_GROUP_ORIGIN == "samples_bcchr",
           hgncSymbol %in% fraser_sample_hits$hgncSymbol) %>%
    dplyr::select(sampleID, hgncSymbol, padjust) %>%
    # parse sample IDs into family and individual codes
    mutate(family_id = map_chr(sampleID,     ~ str_match(., "(DH[0-9]{4})([0-9]{2}|HC)")[2]),
           individual_id = map_chr(sampleID, ~ str_match(., "(DH[0-9]{4})([0-9]{2}|HC)")[3])) %>%
    # mutate(log_padjust = -log(1 + padjust)) %>%
    dplyr::select(-sampleID) %>%
    pivot_wider(names_from = individual_id, values_from = padjust)

plot_family_hits <- function(hit_table, this_family_id, proband_only = FALSE) {
    hit_matrix <-
        hit_table %>%
        filter(family_id == this_family_id) %>%
        { if (proband_only) filter(., !is.na(`01`),) else . } %>%
        column_to_rownames("hgncSymbol") %>%
        dplyr::select(sort(names(.)), -family_id) %>%
        as.matrix() %>%
        -log(.)
    
    ht <- Heatmap(hit_matrix,
            name = paste0(this_family_id, ",\n-log(adj_p-val)"),
            col = viridis::viridis(100),
            cluster_columns = F,
            cluster_rows = F,
            na_col = "grey90",
            width = ncol(hit_matrix) * unit(10, "mm"),
            height = nrow(hit_matrix) * unit(10, "mm"),
            rect_gp = gpar(col = "white", lwd = 4))
    
    ht.draw = draw(ht)
    
    png(file = here::here("plots", paste0(this_family_id, "_hit_summary.png")),
        width = ComplexHeatmap:::width(ht.draw) %>%
            convertX("inch", valueOnly = T),
        height = ComplexHeatmap:::height(ht.draw) %>%
            convertY("inch", valueOnly = T),
        units = "in", res = 300)
    
    draw(ht)
    
    dev.off()
}


walk(unique(fraser_results_sample_hits_table$family_id),
     ~ plot_family_hits(fraser_results_sample_hits_table, ., proband_only = T))

```

## Missing Hit Investigation

It looks like true hits are being dropped because there are occurrences in the GTEx control. I want to determine what the average frequency of hits are within the GTEx dataset and see if a more reasonable threshold can be identified and used.

### Determine hit distribution in GTEx

```{r gtex_hit_distribution}

ggplot(fraser_results_by_gene.tib %>%
           pivot_longer(-hgncSymbol, names_to = "sample_group", values_to = "hit_count")) +
    geom_histogram(aes(x = hit_count), bins = 21) +
    xlim(c(0, 20)) +
    facet_grid(rows = vars(sample_group))

```
